### 2022-05-13 MSC-GUDMUNDSSON-2022.tex
Complete content-wise. Waiting on final feedback.
Small fixes

%% Bachelors: primary advisor (Umsjónarkennari), ONLY ONE!
%% All others: As many as you want
\supervisors{
  \personinfo{Dr. Yngvi Björnsson, PhD}{\MLSupervisor}{\MLProfessor}{\MLReykjavikUniversity}{\MLIceland}
  \personinfo{Dr. Ólafur Rögnvaldsson, PhD}{Examiner}{Managing Director}{Belgingur ehf.}{\MLIceland}
%  \personinfo{Ian M. Great}{Co-advisor}{\MLProfessor}{Hochschule Düsseldorf}{Germany}
}

%% Bachelors: secondary advisor (Leiðbeinandi), ONLY ONE
%% All others: As many as you want
\examiners{
  \personinfo{Tough E. Questions}{\MLExaminer}{Associate Professor}{Massachusetts Institute of Technology}{USA}

@@

\ifdraft{\coverchapter{Important!!!  Read the Instructions!!!} If you
  have not already done so, \LaTeX{} the \path{instructions.tex} to
  learn how to setup your document and use some of the features.  You
  can see a (somewhat recent) rendered PDF of the instructions included in this folder at \path{instructions-publish.pdf}.
  There is also more information on working with \LaTeX{} at
  \url{http://samvinna.ru.is/project/htgaru/how-to-get-around-projects-publish.pdf}.
  This includes common problems and fixes.

  This page will disappear in anything other than draft mode.}{}

@@

\coverchapter{Preface}
% RUM: "Preface (optional)"
This dissertation is original work by the author, Einar~Helgi~Guðmundsson.

\draftnote{The preface is an optional element
  explaining a little who performed what work.  See
  \url{https://www.grad.ubc.ca/sites/default/files/materials/thesis_sample_prefaces.pdf}
  for suggestions.

  List of publications as part of the preface is
  optional unless elements of the work have already been published.
  It should be a comprehensive list of all publications in which
  material in the thesis has appeared, preferably with references to
  sections as appropriate.  This is also a good place to state
  contribution of student and contribution of others to the work
  represented in the thesis.}

>>>>

%% Bachelors: primary advisor (Umsjónarkennari), ONLY ONE!
%% All others: As many as you want
\supervisors{
  \personinfo{Yngvi
  Björnsson}{\MLSupervisor}{\MLProfessor}{\MLReykjavikUniversity}{\MLIceland}

%  \personinfo{Ian M. Great}{Co-advisor}{\MLProfessor}{Hochschule Düsseldorf}{Germany}
}

%% Bachelors: secondary advisor (Leiðbeinandi), ONLY ONE
%% All others: As many as you want
\examiners{
  \personinfo{Ólafur Rögnvaldsson}{\MLExaminer}{Managing Director}{Belgingur ehf.}{\MLIceland}
  \personinfo{Stephan Schiffel}{\MLExaminer}{Assistant Professor}{\MLReykjavikUniversity}{\MLIceland}

@@

@@

### 2022-05-13 background.tex
Complete content-wise. Waiting on final feedback.
Small fixes

\section{Transformers\label{cha:transformer}}
%Hard to train
%sequential
% -> parallelization hard
%encoder/decoder instead. Attention. 
% Explain enc/dec
% Autoencoders generate same output as the input, so no labeled data is needed -> completely unsupervisd earning.
% Locality a problem with CNNs and RNNs. Can easily find local context but keeping on to context far apart in input is difficult. Enter transformers.
 %Encoder reads blocks of data and uses that and previously generated data as inputs into decoder which generatesdata one point at a time.
 %Attention. How you relate things to each other. i.e. input to output
 %self attention. How do features in the input data relate to each other

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/encoder_decoder.jpeg}
    \caption{The encoder/decoder structure \cite{nechu_what_2020}
    \label{fig:encoder_decoder}}
\end{figure}

@@

\section{Weather forecasts\label{cha:WRF}}
Numerical Weather Prediction (NWP) has used mathematical models of the oceans and atmosphere since the 1920s to predict the weather based on current conditions, although realistic results weren't achieved until the advent of computer simulations in the 1950s \cite{noauthor_numerical_2022}. The Weather Research and Forecasting (WRF) model is an NWP system created by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administration (NOAA), the Air Force Weather Agency (AFWA), the Naval Research Laboratory (NRL), the University of Oklahoma (OU), and the Federal Aviation Administration (FAA) in the United States for both research and operational weather forecasting \cite{noauthor_weather_2022}.\\
WRF-SOLAR is an NWP model based on WRF specifically designed to model values useful for solar power, including high frequency irradiance calculations, more accurate solar position algorithms, and more robust aerosol and particle simulations with regard to radiation \cite{jimenez_wrf-solar_2016}.
\section{Previous work}
Lin et al. \cite{lin_temporal_2020} showed that while the more traditional GRUs and LSTMs show good performance on the task, Temporal Convolutional Neural Networks (TCNN) can show even better performance. \textit{\endquote{TCNN is a novel convolutional architecture designed for sequential modelling, which combines causal and dilated convolutions
and residual connections}} \cite{lin_temporal_2020}. \\
Jaidee et al. \cite{jaidee_very_2019} showed that LSTMs, GRUs and derivative methods show relatively similar performance for very short-term predictions on the time scale of a few hours. This timescale is where it is critical to accurately predict cloud movements. 
These methods have been taken as far as building genetic algorithms to generate the optimal neural network for the purpose \cite{jaidee_very_2019}. These genetic algorithms operate on the same principles as animal breeding. They generate multiple candidate neural networks and use the best networks as a base to generate new networks for the next generation. This is done for multiple generations and in the end, the best neural network generated is used \cite{jaidee_very_2019}.\\
Su et al. \cite{su_machine_2019} did extensive experimentation on more conventional neural networks and newer, less known networks including  Non-linear Auto Regressive Neural Networks (NARXNN), as well as various statistical approaches. When predicting power output, NARXNN takes in the current power output values and considers them, in real time, as well as the past values of the power output of the system. These experiments showed that NARXNN had the best performance of these, by a rather large margin, and a hybrid method, of the better performing methods, showed even better results \cite{anderson_using_2018}. The statistical methods generally performed significantly worse in predicting the power output. \\
Satellite imagery has been used for predicting solar power output as well. A neural network was trained to predict solar power production, primarily from cloud cover information it learned from these satellite images \cite{jang_solar_2016}.\\
\textcolor{red}{AlexNet}

>>>>

\section{Transformers\label{cha:transformer}}

%Hard to train
%sequential
% -> parallelization hard
%encoder/decoder instead. Attention. 
% Explain enc/dec
% Autoencoders generate same output as the input, so no labeled data is needed -> completely unsupervisd earning.
% Locality a problem with CNNs and RNNs. Can easily find local context but keeping on to context far apart in input is difficult. Enter transformers.
 %Encoder reads blocks of data and uses that and previously generated data as inputs into decoder which generatesdata one point at a time.
 %Attention. How you relate things to each other. i.e. input to output
 %self attention. How do features in the input data relate to each other

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{imgs/encoder_decoder.jpeg}
    \caption{The encoder/decoder structure \cite{nechu_what_2020}
    \label{fig:encoder_decoder}}
\end{figure}

@@

\section{Weather forecasts\label{cha:WRF}}
Numerical Weather Prediction (NWP) has used mathematical models of the oceans and atmosphere since the 1920s to predict the weather based on current conditions, although realistic results weren't achieved until the advent of computer simulations in the 1950s \cite{noauthor_numerical_2022}. The Weather Research and Forecasting (WRF) model is an NWP system created by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administration (NOAA), the Air Force Weather Agency (AFWA), the Naval Research Laboratory (NRL), the University of Oklahoma (OU), and the Federal Aviation Administration (FAA) in the United States for both research and operational weather forecasting \cite{noauthor_weather_2022}.\\
WRF-SOLAR is an NWP model based on WRF specifically designed to model values useful for solar power, including high frequency irradiance calculations, more accurate solar position algorithms, and more robust aerosol and particle simulations with regard to radiation \cite{jimenez_wrf-solar_2016}.
\section{Previous work}
Lin et al. \cite{lin_temporal_2020} showed that while the more traditional GRUs and LSTMs show good performance on solar power prediction, Temporal Convolutional Neural Networks (TCNN) can show even better performance. \textit{\endquote{TCNN is a novel convolutional architecture designed for sequential modelling, which combines causal and dilated convolutions
and residual connections}} \cite{lin_temporal_2020}. \\
Jaidee et al. \cite{jaidee_very_2019} showed that LSTMs, GRUs and derivative methods show relatively similar performance for very short-term predictions of solar power production on the time scale of a few hours. This timescale is where it is critical to accurately predict cloud movements. 
These methods have been taken as far as building genetic algorithms to generate the optimal neural network for the purpose \cite{jaidee_very_2019}. These genetic algorithms operate on the same principles as animal breeding. They generate multiple candidate neural networks and use the best networks as a base to generate new networks for the next generation. This is done for multiple generations and in the end, the best neural network generated is used \cite{jaidee_very_2019}.\\
Su et al. \cite{su_machine_2019} did extensive experimentation on more conventional neural networks and newer, less known networks including  Non-linear Auto Regressive Neural Networks (NARXNN), as well as various statistical approaches. When predicting solar power output, NARXNN takes in the current power output values and considers them, in real time, as well as the past values of the power output of the system. These experiments showed that NARXNN had the best performance of these, by a rather large margin, and a hybrid method, of the better performing methods, showed even better results \cite{anderson_using_2018}. The statistical methods generally performed significantly worse in predicting the power output. \\
Satellite imagery has been used for predicting solar power output as well. A neural network was trained to predict solar power production, primarily from cloud cover information it learned from these satellite images \cite{jang_solar_2016}.\\

### 2022-05-13 methods.tex
Small fixes

    \noindent\textbf{Gating mechanisms} identify and skip unused input data, granting models greater ability to work with less curated input data, or datasets where it is unknown which parts are of significance.\\
    The Gated Residual Network (GRN) structure, shown in Figure~\ref{fig:ttgt_detail}, is a base building block of TFTs \cite{lim_temporal_2020}. It provides the model the flexibility to choose the complexity of its input, more complex where needed, and less complex where not.\\
    \textbf{Variable selection networks} emphasize the most relevant inputs for every time step \cite{lim_temporal_2020}. Some variables may be more important than others, \enquote{TFT is designed to provide instance-wise variable selection through the use of variable selection networks applied to both static covariates and time-dependent covariates.} \cite{lim_temporal_2020} The variable selection mechanism also allows the model to remove noisy inputs that might have a negative effect on performance. \\
    \textbf{Static covariate encoders} encode static features into context vectors, integrating them into the network \cite{lim_temporal_2020}.  \\
    \textbf{Temporal processing} using a more traditional RNN sequence-to-sequence layer (see Section~\ref{cha:RNN}) for short term, local processing and the more novel multi-headed attention layer (see Section~\ref{cha:transformer}) for long term pattern capture \cite{lim_temporal_2020}.\\
    \textbf{Prediction intervals} in the form of quantile forecasts, discussed in Section~\ref{sec:loss_function}, display a confidence interval for the predictions \cite{lim_temporal_2020}. \\
    Figure~\ref{fig:ttgt_detail} shows a detailed view of the TFT structure as discussed.

@@

    \begin{figure}[ht!]
        \centering
        \includegraphics[scale=0.6]{imgs/solar_TFT_overview.jpg}
        \caption{An overview of the setup of the neural network.
        \label{fig:solar_tft_overviwew}}
    \end{figure}

@@

    \subsubsection{Tuning}
        Significant tuning was done to optimize the performance of both models. Since the goal of tuning was not only to minimize Mean Absolute Percentage Error but balance that with the ability to recognize uncertainty, the process was very manual. Tables~\ref{tab:full_params} and \ref{tab:lim_params} show the selected hyperparameters for the full and limited models respectively.\\

@@

    \subsubsection{Loss function}\label{sec:loss_function}
    The loss function used for Temporal Fusion Transformers is quantile loss, as it's an integral part of generating the confidence intervals of the model \cite{lim_temporal_2020}.
    A model's prediction is a random value within a probability distribution \cite{wen_multi-horizon_2018}. If we make multiple predictions we can estimate this probability distribution by seeing within what range a selected proportion of the predictions fall \cite{koenker_quantile_2001}. PyTorch Forecasting's quantile loss function takes the Mean Absolute Error (MAE) of three quantiles, 90\%, 50\% and 10\%. This gives a good picture of the model's confidence in a prediction. A smaller quantile range indicates more confidence.

>>>>

    \noindent\textbf{Gating mechanisms} identify and skip unused input data, granting models greater ability to work with less curated input data, or datasets where it is unknown which parts are of significance.\\
    The Gated Residual Network (GRN) structure, shown in Figure~\ref{fig:ttgt_detail}, is a base building block of TFTs \cite{lim_temporal_2020}. It provides the model the flexibility to choose the complexity of its input, more complex where needed, and less complex where not.\\
    \clearpage
    \noindent\textbf{Variable selection networks} emphasize the most relevant inputs for every time step \cite{lim_temporal_2020}. Some variables may be more important than others, \enquote{TFT is designed to provide instance-wise variable selection through the use of variable selection networks applied to both static covariates and time-dependent covariates.} \cite{lim_temporal_2020} The variable selection mechanism also allows the model to remove noisy inputs that might have a negative effect on performance. \\
    \textbf{Static covariate encoders} encode static features into context vectors, integrating them into the network \cite{lim_temporal_2020}.  \\
    \textbf{Temporal processing} using a more traditional RNN sequence-to-sequence layer (see Section~\ref{cha:RNN}) for short term, local processing and the more novel multi-headed attention layer (see Section~\ref{cha:transformer}) for long term pattern capture \cite{lim_temporal_2020}.\\
    \textbf{Prediction intervals} in the form of quantile forecasts, discussed in Section~\ref{sec:loss_function}, display a confidence interval for the predictions \cite{lim_temporal_2020}. \\
    Figure~\ref{fig:ttgt_detail} shows a detailed view of the TFT structure as discussed.

@@

    \afterpage{%
    \begin{figure}
        \centering
        \makebox[\textwidth][c]{\includegraphics[scale=0.6]{imgs/solar_TFT_overview.jpg}}
        \caption{An overview of the setup of the neural network.
        \label{fig:solar_tft_overviwew}}
    \end{figure}
    \clearpage
    }

@@

    \pagebreak
    \subsubsection{Tuning}
        Significant tuning was done to optimize the performance of both models. Since the goal of tuning was not only to minimize Mean Absolute Percentage Error but balance that with the ability to recognize uncertainty, the process was very manual. Tables~\ref{tab:full_params} and \ref{tab:lim_params} show the selected hyperparameters for the full and limited models respectively.\\

@@

    \pagebreak
    \subsubsection{Loss function}\label{sec:loss_function}
    The loss function used for Temporal Fusion Transformers is quantile loss, as it's an integral part of generating the confidence intervals of the model \cite{lim_temporal_2020}.
    A model's prediction is a random value within a probability distribution \cite{wen_multi-horizon_2018}. If we make multiple predictions we can estimate this probability distribution by seeing within what range a selected proportion of the predictions fall \cite{koenker_quantile_2001}. PyTorch Forecasting's quantile loss function takes the Mean Absolute Error (MAE) of three quantiles, 90\%, 50\% and 10\%. This gives a good picture of the model's confidence in a prediction. A smaller quantile range indicates more confidence.
